Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-22 paper.qmd
Added absolute error plot
with different UMAP parameter choices

@@

# df_all_mnist <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), umap_data_with_hb_id_pbmc)
# 
# show_langevitour(df_all_pbmc, df_bin_pbmc, df_bin_centroids_pbmc, benchmark_pbmc, distance_pbmc, "distance", col_start = "PC_")

@@

error_df <- error_df |>
  dplyr::mutate(type = paste0(n_neighbors, "_", min_dist))

abs_error_plot_pbmc <- ggplot(error_df, aes(x = num_non_empty_bins,
                                       y = error, group = type,
colour = type)) +
  geom_point() +
  geom_line() +
   # geom_vline(xintercept = 588, linetype="solid",
   #              color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  ylab("absolute error") +
  xlab("number of non-empty bins")

@@

>>>>

@chen2023

@@

# df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), umap_data_with_hb_id_pbmc)
# 
# show_langevitour(df_all_pbmc, df_bin_pbmc, df_bin_centroids_pbmc, benchmark_pbmc, distance_pbmc, "distance", col_start = "PC_")

@@

error_df <- error_df |>
  dplyr::mutate(type = paste0("n_neighbors: ", n_neighbors, ", min_dist: ", min_dist))

abs_error_plot_pbmc <- ggplot(error_df, aes(x = num_non_empty_bins,
                                       y = log(error), group = type,
colour = type)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 278, linetype="solid",
                color = "black", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7),
        legend.text = element_text(size = 3)) +
  scale_color_manual(values=c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")

```{r}
#| echo: false
#| fig-cap: "Absolute error from UMAP applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to 278, the lowest error occurred with the parameters n_neighbors: 5 and min_dist: 0.3."
#| label: fig-abserrorPBMC
#| fig-pos: H
#| out-width: 100%

@@

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.3.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, 
                                                                 x = "UMAP1", y = "UMAP2"))) |>
  dplyr::mutate(cell_label = umap_pbmc$cell_label) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 5, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?"
#| label: fig-umapsuggest
#| fig-pos: H

<!--Fit the model-->

## Decide by looking at MSE plot
num_bins_x_pbmc <- 21
num_bins_y_pbmc <- 53
hex_size_pbmc <- 0.03

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  geom_point(data = umap_pbmc_scaled, aes(x = UMAP1, y = UMAP2), alpha = 0.05) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

# df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), umap_data_with_hb_id_pbmc)
# 
# show_langevitour(df_all_pbmc, df_bin_pbmc, df_bin_centroids_pbmc, benchmark_pbmc, distance_pbmc, "distance", col_start = "PC_")

error_df <- error_df |>
  mutate(type = if_else(total == 0, "no error",
                        if_else(total <= 5, "error 0-5",
                                if_else(total <= 10, "error 5-10",
                                        if_else(total <= 15, "error 10-15",
                                                if_else(total <= 20, "error 15-20",
                                                        "error greter than 20"))))))

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 0-5", "error 5-10", "error 10-15", "error 15-20", "error greter than 20"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=rainbow_hcl(6)) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

```{r}
#| echo: false
#| fig-cap: "(a) Model in the 2D space overlaid on UMAP data, and (b) Error occurred by the high-D model for high-D data and the error mapped into 2D NLDR points."
#| label: fig-nldervisPBMCUMAP2
#| fig-pos: H
